---
title: "Modelling Wetland Loss in Saskatchewan"
format: html
---

## Introduction

The purpose of this report is to demonstrate the modelled loss of wetlands in Saskatchewan,
based on GIS data collected by the CWI. This will then be compared to drainage data collected by WSA.

## Data Exploration

Let us first explore the data. Here we will focus on the data generated by the CWI and stratified by 
virtual basin. Let's take a look at some histograms and potential relationships to investigate.

### Distribution of Drain Lengths and Numbers of Drains
```{r}
#| echo: false
targets::tar_load(vb_db)
```
Let's first take a look at the distribution of drain lengths and the distribution of the number of drains. 
Note here that I am only looking at virtual basins where the drain lengths are greater than 0, or the 
number of drains are greater than 0. That is, we aren't considering undrained virtual basins. This 
also gets rid of incomplete data.

We will also look to see if there is a relationship between number of drains and the length of drains 
per virtual basin. We would likely expect to see a positive relationship (i.e., more drains implies 
more length of drain per virtual basin).
 
```{r}
#| echo: false
targets::tar_load(combined_drains_plot)
print(combined_drains_plot)
```

We can see a fairly clear relationship between the number of drains and the length of drains per virtual basin. 
In fact, we find a correlation of `r cor(vb_db$sum_Length, vb_db$Polyline_C)`. With this in 
mind, we can probabl be confident that we can use either of these two variables in a model as they are effectively 
telling the same story.

### Distribution of Drained Area by CWI Class

Let us now take a look at the distribution of drained area by CWI class. For this, I will present the distribution 
of percent drained virtual basin for classes 1 and 5. This is because the virtual basins are of different sizes, 
and so we should expect to see more drainage in larger basins. So, taking the percent drained per basin 
standardizes the drainage.

Here are the percent drainage for CWI classes 1 and 5:
```{r}
#| echo: false
targets::tar_load(combined_class_drainage_plot)
print(combined_class_drainage_plot)
```

Across the virtual basins for which we have data, we find a total drained area of 
`r round(sum(vb_db$CWI_1, na.rm = TRUE) + sum(vb_db$CWI_5, na.rm = TRUE), 2)`km^2, which comprises a total of 
`r round(((sum(vb_db$CWI_1, na.rm = TRUE) + sum(vb_db$CWI_5, na.rm = TRUE)) / sum(vb_db$CWI_Total, na.rm = TRUE)) * 100 ,2)`%.

### Does number of drains correlate with drainage?

One thing that we might be interested in is whether the number of drains in a virtual 
basin correlates with the drainage in that basin. We would expect this to be the case,
in that we should see a positive relationship. If so, we may be able to use this relationship
in a larger model to estimate how much drainage we are missing in each virtual basin.

Let us plot this relationship:


```{r}
#| echo: false
targets::tar_load(drains_vs_drainage_plot)
print(drains_vs_drainage_plot)
```

### Does landcover class explain variation in drainage?

Let us now have a look at the variation in both number of drains and drained area as it pertains to landcover class. We will use boxplots for this:

```{r}
#| echo: false
targets::tar_load(combined_boxplots)
print(combined_boxplots)
```

### Spatial Distribution of Drainage

Let's take a look at how the drainage varies in space. This will allow us to see if there are any hotspots (or coldspots) of drainage that are not
necessarily influenced by the number of drainage ditches. 

First, let us see the spatial distribution of the number of drains per square kilometer in each VB, recalling
that the number of drains can be used as a proxy for length of drains. We will plot this side by side
with the percent drained wetlands at each virtual basin (grey areas denote NA):

```{r}
#| echo: false
targets::tar_load(combined_spatial_plot)
print(combined_spatial_plot)
```

### Exploration of the WSA Data

Let's now take a look at the data that has been used in the WSA analysis for wetland drainage amount. This is for the objective looking 
at “what does WSA think current intactness levels are, what might actual levels be, and what does that mean for potential to miss the 
retention targets in the future”. Note that for this purpose, the CWI impacts we lump into “intact” or “drained” categories will be 
different than for the other exercise. Here we will consider CWI classes 1, 4, and 5 to be drained and 0, 2, and 3 to be intact.

Because we are interested in current intactness levels according to WSA, let us examine the % intactness for each planning unit in the following table:

```{r}
#| echo: false
targets::tar_load(wsa_data_wide)
wsa_data_wide$Percent_Intact <- round((rowSums(wsa_data_wide[, c(3,5,7)], na.rm = TRUE) / rowSums(wsa_data_wide[,3:8], na.rm = TRUE)) * 100, 2)
print(wsa_data_wide[, c("Name", "Percent_Intact")])
```

## Modelling Drainage to Quantify Uncertainty

Let us now attempt to create some models! TO DO INSERT ACTUAL MODEL HERE.

Here is the Stan summary for the Null model:

```{r}
#| echo: false
targets::tar_load(model_summary_null)
print(model_summary_null) 
```

And now here is the Stan summary for the model that relates number of drains to drainage:

```{r}
#| echo: false
targets::tar_load(model_summary_n_drains)
print(model_summary_n_drains)
```

Stan summary for the model that relates drainage to WSA, but shares a variance across WSAs:
```{r}
#| echo: false
targets::tar_load(model_summary_wsa)
print(model_summary_wsa)
```
